---
title: "DATA 606 Project Proposal"
author: "Sie Siong Wong"
date: "10/13/2019"
output: html_document
---

### Data Preparation

```{r, setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

# Load the packages required to tidy and transform the data.

library(dplyr)
library(tidyr)
library(reshape2)
library(stringr)
library(psych)
library(ggplot2)


# load data

GDP <- read.csv("https://raw.githubusercontent.com/SieSiongWong/DATA-606/master/GDP.csv", header=TRUE, sep=",")

Crime <- read.csv("https://raw.githubusercontent.com/SieSiongWong/DATA-606/master/Crime%20Rate.csv", header=TRUE, sep=",")

Unemployment <-  read.csv("https://raw.githubusercontent.com/SieSiongWong/DATA-606/master/Unemployment%20Rate.csv", header=TRUE, sep=",")

GDPvsCrime <- read.csv("https://raw.githubusercontent.com/SieSiongWong/DATA-606/master/GDPvsCrimebyYear.csv", header=TRUE, sep=",")

# Clean and reshape the GDP data.

GDP <- GDP %>% rename("States"="X")  # Change column name.

GDP <- GDP %>% melt(GDP, id.vars=c("States"), measure.vars=2:ncol(GDP), variable.name="Year", value.name="GDP", na.rm=TRUE) %>% mutate(Year = as.numeric(gsub("X", "", Year))) # Turn into long form. 

# Clean and reshape the Crime data.

Crime <- Crime %>% rename("Year"="X") # Change column name.

total_col <- apply(Crime[,-1], 1, sum)      # Change to percentage rate.
Crime2 <- lapply(Crime[,-1], function(x) {
  x / total_col*100
})

Crime2$Year <- Crime$Year # Merge two data frames.
Crime2 <- merge(Crime2, Crime, by="Year")
Crime2 <- Crime2[,-c(52:101)]


Crime2 <- Crime2 %>% melt(Crime2, id.vars=c("Year"), measure.vars=2:ncol(Crime2), variable.name="States", value.name="CrimeRate", na.rm=TRUE)  # Turn into long form.       

Crime2$States <- sub("\\.x$","", Crime2$States)
Crime2$States <- sub("\\."," ", Crime2$States) # Remove dot in states name.


Crime2 <- dcast(Crime2, States~Year, value.var="CrimeRate") # dcast the dataset into wide form.

Crime2 <- Crime2 %>% melt(Crime2, id.vars=c("States"), measure.vars=2:ncol(Crime2), variable.name="Year", value.name="CrimeRate", na.rm=TRUE)  # Turn into long form again to make it consistent with other twos.  

# Clean and reshape the Unemployment data.

Unemployment <- Unemployment %>% rename("States"="X")  # Change column name.

Unemployment <- Unemployment %>% melt(Unemployment, id.vars=c("States"), measure.vars=2:ncol(Unemployment), variable.name="Year", value.name="UnemploymentRate", na.rm=TRUE) %>% mutate(Year = as.numeric(gsub("X", "", Year))) # Turn into long form. 

# Join the datasets into single dataset.

Merged_df <- merge(GDP,Unemployment, by=c("States", "Year"))
Merged_df <- merge(Merged_df, Crime2, by=c("States", "Year"))

```


### Research question 

**Is better GDP will contribute to lower unemployment rate and/or lower violent crime rate across the states or the nation as a whole?**


### Cases 

**Each case represents a GDP, crime rat, or unemployment rate for each state. There are total of 3 datasets: GDP, Crime, Unemployment for all 50 states from year 1997 to 2014. Therefore, each dataset will have 900 observations and 2700 observations in total for the 3 datasets.**


### Data collection 

**Annual GDP Data: Data is collected and stored by Bureau of Economic Analysis.**

**Annual Unemployment Rate Data: Data is collected and stored by U.S. Bureau of Labor Statistics.**

**Annual Crime Rate Data: Data is collected and stored by Uniform Crime Reporting Statistics.**


### Type of study 

**This is an observational study.**


### Data Source 

**Annual GDP Data: Data is collected and stored by BEA and is available online here: https://apps.bea.gov/iTable/index_regional.cfm. The data was extracted using the BEA's interactive data table and saved to a csv file to be used for this project.**

**Bureau of Economic Analysis. U.S. Department of Commerce. Retrieved [10/13/2019] from https://apps.bea.gov/iTable/index_regional.cfm.**

**Annual Unemployment Rate Data: Data is collected and stored by Iowa Community Indicators Program from the BLS and is available here: https://www.icip.iastate.edu/tables/employment/unemployment-states. The data was in a excel file and was last updated on April 2016 and ready to be used for this project.**

**Annual Unemployment Rates by State. (April 2016). Iowa Community Indicators Program . Retrieved from https://www.icip.iastate.edu/tables/employment/unemployment-states.**

**Annual Crime Rate Data: Data is collected and stored by UCR and is available here: https://www.ucrdatatool.gov/Search/Crime/State/TrendsInOneVar.cfm?NoVariables=Y&CFID=188098989&CFTOKEN=63a6599343a03796-EA1C8CE0-D66E-C7C2-ABDE73D498B77930. The data was extracted using the site's Get Table tool and saved to a csv file to be used for this project.**

**Uniform Crime Reporting Statistics. Federal Bureau of Investigation. U.S. Department of Justice. Retrieved [10/13/2019] from https://www.ucrdatatool.gov/Search/Crime/State/TrendsInOneVar.cfm?NoVariables=Y&CFID=188098989&CFTOKEN=63a6599343a03796-EA1C8CE0-D66E-C7C2-ABDE73D498B77930.**



### Dependent Variable

**The response variables are unemployment rate and crime rate and are numerical.**



### Independent Variable

**The explanatory variable is GDP and is numerical.**



### Relevant summary statistics 


```{r, eval=TRUE}

## Average Annual Unemployment Rate Distribution between year 1997 and 2014.

# Summary statistics for the unemployment rate variable.
describe(Unemployment$UnemploymentRate)

# Average annual unemployment rate for each state.
UnemploymentRate_Mean <- Unemployment %>% group_by(States) %>% summarize(Average=round(mean(UnemploymentRate), digits=2))

# Plot a histogram to show the distribution of the average annual unemployment rate.
hist(UnemploymentRate_Mean$Average, main="Average Annual Unemployment Rate Distribution", xlab="Mean", ylab="Frequency", ylim=c(0,12), xlim=c(2.5,8), col="hotpink", breaks=10)

# Plot a normal Q-Q Plot to further show that the distribution of the average annual unemployment rate is close to normal distribution.
qqnorm(UnemploymentRate_Mean$Average)
qqline(UnemploymentRate_Mean$Average)

# Plot a boxplot to show the variation of the unemployment rate across 50 states from year 1997 to 2014.
ggplot(Unemployment, aes(x=reorder(factor(States), UnemploymentRate, fun=median),y=UnemploymentRate,fill=factor(States))) + geom_boxplot() + labs(title="Unemployment Rate by States") + ylab("%") + theme(legend.position = "none", axis.title.x = element_blank(), axis.text.x=element_text(angle=90)) + theme(plot.title = element_text(hjust=0.5)) + theme(axis.text.x = element_text(margin = margin(t = 25, r = 20, b = 0, l = 0)))



## Average Annual GDP Distribution between year 1997 and 2014.

# Summary statistics for the GDP variable.
describe(GDP$GDP)

# Average annual GDP for each state.
GDP_Mean <- GDP %>% group_by(States) %>% summarize(Average=round(mean(GDP), digits=2))

# Plot a histogram to show the distribution of the average annual GDP.
hist(GDP_Mean$Average, main="Average Annual GDP Distribution", xlab="Mean", ylab="Frequency", ylim=c(0,25), col="hotpink", breaks=10)

# Plot a normal Q-Q Plot to further show that the distribution of the average annual GDP is right skewed.
qqnorm(GDP_Mean$Average)
qqline(GDP_Mean$Average)

# Plot a boxplot to show the variation of the GDP across 50 states from year 1997 to 2014.
ggplot(GDP, aes(x=reorder(factor(States), GDP, fun=median),y=GDP,fill=factor(States))) + geom_boxplot() + labs(title="GDP by States") + ylab("%") + theme(legend.position = "none", axis.title.x = element_blank(), axis.text.x=element_text(angle=90)) + theme(plot.title = element_text(hjust=0.5)) + theme(axis.text.x = element_text(margin = margin(t = 25, r = 20, b = 0, l = 0)))


## Average Annual Crime Rate Distribution between year 1997 and 2014.

# Summary statistics for the crime rate variable.
describe(Crime2$CrimeRate)

# Average annual crime rate for each state.
Crime2_Mean <- Crime2 %>% group_by(States) %>% summarize(Average=round(mean(CrimeRate), digits=2))

# Plot a histogram to show the distribution of the average annual crime rate.
hist(Crime2_Mean$Average, main="Average Annual Crime Rate Distribution", xlab="Mean", ylab="Frequency", ylim=c(0,25), col="hotpink", breaks=10)

# Plot a normal Q-Q Plot to further show that the distribution of the average annual crime rate is right skewed.
qqnorm(GDP_Mean$Average)
qqline(GDP_Mean$Average)

# Plot a boxplot to show the variation of the crime rate across 50 states from year 1997 to 2014.
ggplot(Crime2, aes(x=reorder(factor(States), CrimeRate, fun=median),y=CrimeRate,fill=factor(States))) + geom_boxplot() + labs(title="Crime Rate by States") + ylab("%") + theme(legend.position = "none", axis.title.x = element_blank(), axis.text.x=element_text(angle=90)) + theme(plot.title = element_text(hjust=0.5)) + theme(axis.text.x = element_text(margin = margin(t = 25, r = 20, b = 0, l = 0)))


# Plot Average Annual GDP vs Unemployment Rate across states.
Merged_df2 <- Merged_df %>% group_by(States) %>% summarize(Average_GDP=mean(GDP), Average_Unemployment=mean(UnemploymentRate), Average_Crime=mean(CrimeRate))

ggplot(Merged_df2,aes(States)) + ggtitle("GDP vs. Unemployment Rate") + ylab("%") +  geom_line(aes(y=Average_GDP, group=1, colour="GDP")) + geom_point(aes(y=Average_GDP, group=1), color="Red") + geom_line(aes(y=Average_Unemployment, group=2, colour="Unemployment")) + geom_point(aes(y=Average_Unemployment, group=2), color="Blue") + theme(plot.title = element_text(hjust=0.5), axis.title.x=element_blank(), axis.text.x=element_text(angle=90,hjust=1), legend.position=c(0.5,0.9),legend.title=element_blank())


# Plot Annual GDP vs Violent Crime Rate in United States.
ggplot(GDPvsCrime,aes(x=Year)) +  geom_line(aes(y=GDPMillions, colour="GDP"))  + geom_line(aes(y=TotalCrime*10, colour="Crime")) + scale_y_continuous(sec.axis = sec_axis(~ ./100 , name = "Total Crime")) + scale_colour_manual(values=c("blue","red")) + labs(y="GDP in Millions", x="Year", colour="Parameter") + theme(legend.position=c(0.5,0.9))

```

